---
title: "haplotype_reconstruction_qtl-nf: Quality Control Report"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    code_folding: hide
date: "`r format(Sys.time(), '%m-%d-%Y')`"
---

```{r setup, include=FALSE}
library(purrr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(knitr)
library(qtl2)
library(broman)
library(qtlcharts)
library(fst)

# project directory
knitr::opts_knit$set(root.dir = getwd())
# knitr::opts_knit$set(root.dir = "/flashscratch/widmas/HR_QC_outputDir/work/95/466eb43ab299946e7f2518a5dcce37")
knitr::opts_chunk$set(echo = TRUE, dev = "png")

options(scipen = 999999)
```

## Load Data

```{r load data}
# cross object
cross <- readRDS(list.files(pattern = "cross.rds"))

# x chromosome intensities
xint <- read.csv(list.files(pattern = "chrX_intensities.csv"), check.names = F)

# y chromosome intensities
yint <- read.csv(list.files(pattern = "chrY_intensities.csv"), check.names = F)

# intensities
int <- fst::read.fst(list.files(pattern = "all_intensities.fst"))

# viterbi
m <- readRDS(list.files(pattern = "maxmarg.rds"))

# viterbi
bad_markers <- readRDS(list.files(pattern = "bad_markers.rds"))

# genoprob errors
e <- readRDS(list.files(pattern = "genotyping_errors.rds"))
```

## Project Metrics

There were `r qtl2::n_ind(cross)` samples in the original experiment, comprising `r nrow(cross$covar[which(cross$covar$sex %in% c("f","F","female")),])` females and `r nrow(cross$covar[which(cross$covar$sex %in% c("m","M","male")),])` males. There were a total of `r sum(qtl2::n_mar(cross))` informative markers among `r length(qtl2::n_typed(cross))` genotyped individuals.

## Marker Metrics

### Missing Genotypes

```{r, missing marker genotypes}
# calculate what percent of samples are missing each marker
percent_missing_marker <- qtl2::n_missing(cross, "marker", "prop")*100

# plot
hist(percent_missing_marker, breaks = 100,
     main = "Distribution of the proportion of samples missing each marker", 
     xlab = "Percent of samples missing genotype",
     ylab = "Number of genotypes")
```

Of all markers, `r length(bad_markers)` markers were missing in at least 10% of samples.

## Sample Metrics

### Missing genotypes

```{r missing genotypes}
# label values with mouse ids
percent_missing_ind_cross <- qtl2::n_missing(cross, "ind", "prop")*100
labels <- paste0(names(percent_missing_ind_cross), " (", round(percent_missing_ind_cross), "%)")

# count number of samples at different missingness thresholds
missing_thresh <- unlist(lapply(c(1,5,10,20), function(x) length(percent_missing_ind_cross[percent_missing_ind_cross > x])))

# plot
qtlcharts::iplot(x = seq_along(percent_missing_ind_cross),
                 y = percent_missing_ind_cross, 
                 indID=labels,
                 chartOpts=list(xlab="Mouse", 
                                ylab="Percent missing genotype data",
                                ylim=c(0, 100))) %>% suppressMessages()

# # view intensities
# int <- int[int$snp %in% qtl2::marker_names(cross),]
# int <- int[seq(1, nrow(int), by=2),-(1:2)] + int[-seq(1, nrow(int), by=2),-(1:2)]
# int <- int[,which(colnames(int) %in% qtl2::ind_ids(cross))]
# n <- names(sort(percent_missing_ind_cross, decreasing=TRUE))
# iboxplot(log10(t(int[,n])+1), orderByMedian=FALSE, chartOpts=list(ylab="log10(SNP intensity + 1)"))
```

There were:

-   `r missing_thresh[1]` samples missing genotypes for more than 1% of markers,
-   `r missing_thresh[2]` samples missing more than 5%,
-   `r missing_thresh[3]` samples missing more than 10%, and
-   `r missing_thresh[4]` samples missing more than 20%.

### Genotyping errors

Statistical support for genotyping errors among samples was calculated from the genotype probabilities.

```{r display the distri}
e <- do.call("cbind", e)
errors_ind <- rowSums(e>2)/n_typed(cross)*100
lab <- paste0(names(errors_ind), " (", round(percent_missing_ind_cross,1), "% missing genotypes)")
iplot(x = seq_along(errors_ind), y = errors_ind, indID=lab,
      chartOpts=list(xlab="Mouse", ylab="Percent genotyping errors", ylim=c(0, 4.1),
                     axispos=list(xtitle=25, ytitle=50, xlabel=5, ylabel=5)))
```

### Sample duplicates

Below is a list of sample pairs flagged as sample duplicates (if there were any). A histogram of the proportion of matching genotypes between all sample pairs is also provided.

```{r sample duplicates}
# Sample Duplicates
cg <- compare_geno(cross, cores=parallel::detectCores()/2)

# default qtl2 report for sample duplicates
summary(cg)

# histogram
hist(cg[upper.tri(cg)], breaks=seq(0, 1, length=201),
     main="", yaxt="n", ylab="", xlab="Proportion matching genotypes")
rug(cg[upper.tri(cg)])
```

### Sex Checks

```{r calculating average sex chr intensities}
# see if there are any samples where a bunch of sex chromosome intensities are missing and remove them?
sex_chr_flag <- list()
for(s in 1:ncol(xint)){
  sex_chr_flag[[s]] <- length(which(is.na(xint[,s])))
}
sex_chr_flag <- colnames(xint[,unlist(sex_chr_flag) > (nrow(xint)*0.85)])
if(length(sex_chr_flag) > 0){
  sex_chr_flag_samples <- sex_chr_flag
  xint <- xint[,!colnames(xint) %in% sex_chr_flag_samples]
  yint <- yint[,!colnames(yint) %in% sex_chr_flag_samples]
}

# average the good X markers
filtered_xint <- xint %>%
  dplyr::filter(!marker %in% names(which(percent_missing_marker > 5)))
xint_ave <- data.frame(colMeans(filtered_xint[complete.cases(filtered_xint),-1])) %>%
  dplyr::mutate(sample = rownames(.))
colnames(xint_ave) <- c("ave_x","sample")
rownames(xint_ave) <- NULL

# average the good Y markers
filtered_yint <- yint %>%
  dplyr::filter(!marker %in% names(which(percent_missing_marker > 5)))
yint_ave <- data.frame(colMeans(filtered_yint[complete.cases(filtered_yint),-1])) %>%
  dplyr::mutate(sample = rownames(.))
colnames(yint_ave) <- c("ave_y","sample")
rownames(yint_ave) <- NULL

# put the Chr X and Y intensities together
ave_int <- dplyr::left_join(xint_ave, yint_ave) %>%
  dplyr::select(sample, everything())
sex_int_covar <- data.frame(cross$covar) %>% 
  dplyr::mutate(sample = rownames(.)) %>%
  dplyr::left_join(ave_int,.) %>%
  dplyr::select(sample, ave_x, ave_y, sex, everything())

# categorize samples into predicted sex chromosome genotypes using intensities
ave_int_predicted <- sex_int_covar %>%
  dplyr::mutate(predicted_sex = dplyr::case_when(ave_x > 0.45 & ave_y < 0.15 ~ "F",
                                             ave_x < 0.45 & ave_y < 0.15 ~ "XO",
                                             ave_x > 0.45 & ave_y > 0.15 ~ "XXY",
                                             ave_x < 0.45 & ave_y > 0.15 ~ "M")) %>%
  dplyr::select(sample, sex, predicted_sex, everything())
percent_missing_ind_df <- data.frame(percent_missing_ind_cross) %>%
  dplyr::mutate(sample = rownames(.)) %>% 
  dplyr::mutate(percent_missing = paste0(round(percent_missing_ind_cross,2),"%"))
rownames(percent_missing_ind_df) <- NULL
sex_check_df <- dplyr::left_join(ave_int_predicted, percent_missing_ind_df) %>%
  dplyr::mutate(assigned_sex = sex)
```

We used the mean intensities of probes from each sex chromosome to infer the genetic sex of each individual.

```{r plot xint by yint}
# make cutoffs df
categories_df <- data.frame(c(0.25,0.5,0.25,0.5),
                            c(0.5,0.5,0.1,0.1),
                            c("M","XXY","XO","F"))
colnames(categories_df) <- c("X","Y","assigned_sex")

# sex palette
pal <- c("green","brown","pink","purple")
names(pal) <- categories_df$assigned_sex

sex_check_plot <- ggplot() +
  geom_jitter(sex_check_df, mapping = aes(x = ave_x, y = ave_y, 
                                     fill = predicted_sex, 
                                     label = assigned_sex,
                                     label2 = percent_missing), 
              size = 3, shape = 21) +
  scale_fill_manual(values = pal) + 
  geom_hline(yintercept = 0.15) + 
  geom_vline(xintercept = 0.45) +
  labs(x = "Mean Chr X Probe Intensity",
       y = "Mean Chr Y Probe Intensity")%>%
  suppressWarnings()

# interactive sex check plot
plotly::ggplotly(sex_check_plot, tooltip = c("assigned_sex","percent_missing"))
```

We used the fraction of heterozygous genotypes called on the X chromosome to obtain additional support.

```{r plot phet by xint}
# proportion of heterozygous X SNPs
pHet <- rowSums(cross$geno$X == 2)/rowSums(cross$geno$X != 0)
pHet_df <- data.frame(pHet) %>%
  dplyr::mutate(sample = rownames(.))
pHet_sexes <- dplyr::left_join(sex_check_df, pHet_df)
pHet_plot <- ggplot() +
  geom_jitter(data = pHet_sexes, mapping = aes(x = ave_x, y = pHet, 
                                  fill = predicted_sex, 
                                  label = assigned_sex,
                                  label2 = percent_missing), 
              size = 3, shape = 21) +
  scale_fill_manual(values = pal) +
  labs(x = "Mean Chr X Probe Intensity",
       y = "Proportion Heterozygous Chr X Genotypes") %>%
  suppressWarnings()

# interactive pHet plot
plotly::ggplotly(pHet_plot, tooltip = c("assigned_sex","percent_missing"))
```

### Crossover Counts

```{r counting crossovers}
# count crossovers for DO ESC samples
GM_xos <- qtl2::count_xo(m, quiet = F, cores = parallel::detectCores()/2)

# make into summarized data frame
if("XO" %in% sex_check_df$provided_sex){
  
  # user specified sex chromosome ploidy as the sex, we can use this information
  
  message("User specified aneuploid individuals in original metadata file; displaying those assignments")
  GM_total_xos <- data.frame(GM_xos, check.names = F) %>%
    dplyr::mutate(sample = rownames(.),
                  sample = gsub("X","",sample)) %>%
    tidyr::pivot_longer(-sample, names_to = "chr", values_to = "xos") %>%
    dplyr::left_join(sex_check_df,.) %>%
    dplyr::mutate(auto = dplyr::if_else(chr == "X", "X", "Autosome"),
                  gen = as.numeric(gen)) %>%
    dplyr::group_by(sample, predicted_sex, sex, assigned_sex, provided_sex, gen, auto) %>%
    dplyr::summarise(total_xos = sum(xos))
  
} else {
  
  GM_total_xos <- data.frame(GM_xos, check.names = F) %>%
    dplyr::mutate(sample = rownames(.),
                  sample = gsub("X","",sample)) %>%
    tidyr::pivot_longer(-sample, names_to = "chr", values_to = "xos") %>%
    dplyr::left_join(sex_check_df,.) %>%
    dplyr::mutate(auto = dplyr::if_else(chr == "X", "X", "Autosome"),
                  gen = as.numeric(gen)) %>%
    dplyr::group_by(sample, predicted_sex, sex, assigned_sex, gen, auto) %>%
    dplyr::summarise(total_xos = sum(xos))
}

```

We can identify additional support for genotype quality using the crossover counts on the autosomes and X chromosome separately

#### Autosomes

```{r plotting autosomal crossovers}

if("XO" %in% GM_total_xos$provided_sex){
  # autosomes
  auto_xo_plot <- ggplot(GM_total_xos %>%
               dplyr::filter(auto == "Autosome")) + 
    theme_bw() + 
    geom_point(mapping = aes(x = gen,
                           y = total_xos,
                           fill = predicted_sex,
                           label = provided_sex),
             size = 2,
             position = position_jitter(width = 0.05)) +
    scale_fill_manual(values = pal, name = "Predicted Sex") +
    scale_colour_manual(values = pal, name = "Predicted Sex") +
    geom_smooth(mapping = aes(x = gen, 
                              y = total_xos, 
                              colour = predicted_sex), 
              method = "lm", se = F) +
    labs(y = "Total crossovers per sample",
         x = "DO Generation")
  plotly::ggplotly(auto_xo_plot, tooltip = c("provided_sex"))
  
  
} else {
  
  # autosomes
  auto_xo_plot <- ggplot(GM_total_xos %>%
               dplyr::filter(auto == "Autosome")) + 
    theme_bw() + 
    geom_point(mapping = aes(x = gen,
                           y = total_xos,
                           fill = predicted_sex,
                           label = assigned_sex),
             size = 2,
             position = position_jitter(width = 0.05)) +
    scale_fill_manual(values = pal, name = "Predicted Sex") +
    scale_colour_manual(values = pal, name = "Predicted Sex") +
    geom_smooth(mapping = aes(x = gen, 
                              y = total_xos, 
                              colour = predicted_sex), 
              method = "lm", se = F) +
    labs(y = "Total crossovers per sample",
         x = "DO Generation")
  plotly::ggplotly(auto_xo_plot, tooltip = c("assigned_sex"))

}

```

#### Chromosome X

```{r plotting chr X crossovers}
if("XO" %in% GM_total_xos$provided_sex){
  
  # sex chromosomes
  sexchr_xo_plot <- ggplot(GM_total_xos %>%
               dplyr::filter(auto == "X")) + 
    theme_bw() + 
    geom_point(mapping = aes(x = gen,
                           y = total_xos,
                           fill = predicted_sex,
                           label = provided_sex),
             size = 2,
             position = position_jitter(width = 0.05)) +
    scale_fill_manual(values = pal, name = "Predicted Sex") +
    scale_colour_manual(values = pal, name = "Predicted Sex") +
    geom_smooth(mapping = aes(x = gen, 
                              y = total_xos, 
                              colour = predicted_sex), 
              method = "lm", se = F) +
    labs(y = "Total crossovers per sample",
         x = "DO Generation")
  plotly::ggplotly(sexchr_xo_plot, tooltip = c("provided_sex"))
} else {
  
  # sex chromosomes
  sexchr_xo_plot <- ggplot(GM_total_xos %>%
               dplyr::filter(auto == "X")) + 
    theme_bw() + 
    geom_point(mapping = aes(x = gen,
                           y = total_xos,
                           fill = predicted_sex,
                           label = assigned_sex),
             size = 2,
             position = position_jitter(width = 0.05)) +
    scale_fill_manual(values = pal, name = "Predicted Sex") +
    scale_colour_manual(values = pal, name = "Predicted Sex") +
    geom_smooth(mapping = aes(x = gen, 
                              y = total_xos, 
                              colour = predicted_sex), 
              method = "lm", se = F) +
    labs(y = "Total crossovers per sample",
         x = "DO Generation")
  
  plotly::ggplotly(sexchr_xo_plot, tooltip = c("assigned_sex"))
}
```

### Sample Status Update

Parsing the proportion of missing genotypes, predicted and recorded sexes, and sample duplication, the following samples have been flagged for follow-up:

```{r sample status update}
# pull samples with duplicates
flagged_pairs <- apply(cg, 2, function(x){
  y <- x[which(x < 1)]
  z <- y[which(y > 0.9)]
  return(z)
})
sample_dups <- lapply(names(unlist(flagged_pairs)), function(x){
  strsplit(x, "[.]")[[1]]
  }) %>% unlist()

if(exists("FLAG_dup_df")){
  rm(FLAG_dup_df)
}
if(!is.null(sample_dups)){
 FLAG_dup_df <- data.frame(sample_dups, "sample duplicate") %>%
  `colnames<-`(c("sample","flag_reason"))
} else {
FLAG_dup_df <- data.frame(NA, "sample duplicate") %>%
  `colnames<-`(c("sample","flag_reason"))
}

# pull samples with many missing genotypes
FLAG_pct_miss <- data.frame(percent_missing_ind_cross[percent_missing_ind_cross > 5]) %>%
  dplyr::mutate(sample = rownames(.)) %>%
  `colnames<-`(c("notes","sample")) %>%
  dplyr::mutate(notes = round(notes, 2),
                notes = paste0(notes,"% of markers missing genotypes"))
if(nrow(FLAG_pct_miss) > 1){
  rownames(FLAG_pct_miss) <- NULL
  FLAG_pct_miss$flag_reason <- "missing genotypes"
}

if(exists("sex_chr_flag_samples")){
  FLAG_sex_chr_int <- data.frame(sex_chr_flag_samples, "sex chromosome intensities missing") %>%
  `colnames<-`(c("sample","flag_reason"))
}



# pull mismatched assigned and predicted sex
if("XO" %in% GM_total_xos$provided_sex){
  FLAG_sex_mismatch_df <- GM_total_xos %>%
    dplyr::ungroup() %>%
    dplyr::distinct(sample, provided_sex, predicted_sex) %>%
    dplyr::mutate(predicted_sex2 = case_when(predicted_sex == "M" ~ "XY",
                                             predicted_sex == "F" ~ "XX")) %>%
    dplyr::filter(provided_sex != predicted_sex2) %>%
    dplyr::mutate(notes = paste0("predicted sex: ",predicted_sex,"; recorded sex: ", provided_sex),
                flag_reason = "sex mismatch") %>%
    dplyr::select(sample, notes, flag_reason)
  rownames(FLAG_sex_mismatch_df) <- NULL
} else {
  FLAG_sex_mismatch_df <- GM_total_xos %>%
    dplyr::ungroup() %>%
    dplyr::distinct(sample, sex, predicted_sex) %>%
    dplyr::mutate(sex = stringr::str_to_upper(sex)) %>%
    dplyr::filter(sex != predicted_sex) %>%
    dplyr::mutate(notes = paste0("predicted sex: ",predicted_sex,"; recorded sex: ", sex),
                  flag_reason = "sex mismatch") %>%
    dplyr::select(sample, notes, flag_reason)
  rownames(FLAG_sex_mismatch_df) <- NULL
}

# pull samples with wild crossover counts
FLAG_xo <- GM_total_xos %>%
  dplyr::ungroup() %>%
  dplyr::distinct(sample, auto, total_xos) %>%
  tidyr::pivot_wider(names_from = auto, values_from = total_xos) %>%
  dplyr::mutate(nxos = Autosome+X) %>%
  dplyr::select(sample, nxos) %>%
  dplyr::filter(nxos > 2000 | nxos < 200) %>%
  dplyr::mutate(notes = paste0(nxos, " crossovers; flagged as too low or high"),
                flag_reason = "crossover counts") %>%
  dplyr::select(-nxos)

# combine all the reports
flag_report <- lapply(ls(pattern = "FLAG"), function(x) get(x)) %>%
  Reduce(dplyr::bind_rows,.) %>%
  dplyr::select(sample, flag_reason, notes) %>%
  dplyr::arrange(sample) %>%
  dplyr::filter(!is.na(sample)) %>%
  dplyr::distinct(sample, flag_reason, notes) %>%
  dplyr::mutate(notes = dplyr::if_else(flag_reason %in% c("sample duplicate","sex chromosome intensities missing"),
                                       true = "T", false = notes))
rownames(flag_report) <- NULL

# tidy the sample flag report?
wide_flag_report <- flag_report %>%
  tidyr::pivot_wider(names_from = flag_reason,
                     values_from = notes)

# display flagged samples
DT::datatable(data = wide_flag_report, 
              class = 'cell-border stripe', 
              options = list(pageLength = 20), 
              rownames = FALSE)

# write flagged samples
write.csv(flag_report, quote = F, row.names = F, file = "sample_QC.csv")
```
